---
title: The effects of job matching on employment accessibility measures in the assessment of core public transport infrastructre
authors:
  - name: J Rafael Verduzco-Torres
    thanks: "Corresponding author. Email: JoseRafael.Verduzco-Torres@glasgow.ac.uk."
    department: Urban Studies
    affiliation: University of Glasgow
    location: Hillhead, Glasgow G12 8RZ, United Kingdom
    email: JoseRafael.Verduzco-Torres@glasgow.ac.uk
  - name: David Philip McArthur
    department: Urban Studies
    affiliation: University of Glasgow
    location: Hillhead, Glasgow G12 8RZ, United Kingdom
    email: David.McArthur@glasgow.ac.uk
abstract: |
 The definition of accessibility emphasizes the attractiveness of opportunities that residents deem valuable. This study revises the common assuption in empirical studies that residents are equally attracted by all type of employment and its implication for public transport evaluation. Additionally, the role of the modifiable problem unit is also examined in this relationship.
keywords:
  - Employment accessibility
  - Job matching
  - Public transport infrastructure
  - Modifiable areal unit problem
bibliography: references.bib
biblio-style: unsrt
output: rticles::arxiv_article
header-includes:
  - \usepackage{float}
  - \usepackage{placeins}
  - \usepackage{amsmath}
  - \usepackage{setspace}
  - \doublespacing
  - \usepackage{graphicx} %Loading the package
  - \graphicspath{{../output/plot/}} %Setting the graphicspath
  - \usepackage{lscape}
---

```{r prelim, include=FALSE}
# Packages
library(tidyverse)
library(modelsummary)
library(kableExtra)

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```


# Introduction

Location-based accessibility measures are key to account for the level of service offered by the transport infrastructure in a variety of studies, including the property market or equity outcomes of large public transport infrastructure projects. An elemental definition of accessibility emphasizes the attractiveness of opportunities that residents deem valuable. This study revises the common assumption in empirical studies that residents are equally attracted by all type of employment. Additionally, the role of the modifiable problem unit is also examined in this relationship. These considerations can offer useful guidance for the environmental measure of accessibility and its application in environmental studies and the distribution of the benefits of public transport infrastructure.

1. What are the key spatial differences between generic and matching accessibility measures and what is the role of the MAUP?
2. How does the evaluation of PT infrastructure differs when accounting for matching employment opportunities in accessibility measures?
3. How do benefits of PT extensions differ between deprived or low-skilled populations and the general population when considering employment matching opportunities in accessibility?

This paper contributes to literature by:

* Examining the spatial distribution of pubic transport benefits using accessibility measures by relaxing the assumption that all employment opportunities are equally attractive to all the population drawing on an empirical case in a ten-year period.
* Examines the role of the MAUP in this effects.
* Reports the findings of GMC, an understudied mega-city, by identifying and digitally reconstructing in a common format (general transit feed specification, GTFS) a series of temporal stages of the MPTN in GMC for a ten-year period (2009-2019).


# Literature review

@Theriault2013 conceptualize accessibility as “the ability that persons living in that area can, within an acceptable travel time, reach a satisfactory diversity of activity locations that are important to them” (p. 234). @Levinson2020 simply put it as "the ease of reaching valued destinations" (p. 130).

## Accessibility measures

<!-- The general location-based formulation consists of the sum of the product of two functions, namely $g(\cdot)$ and $f(\cdot)$, as shown in the following equation [@Paez2012]: -->

<!-- \begin{equation} -->
<!--    A^p_{ik} =  \sum_j g(W_{jk}) f(c^p_{ij}). -->
<!--   \label{eq:location-based1} -->
<!-- \end{equation} -->

<!-- In this expression, accessibility $A$ is viewed from a fixed location $i$ (considered to be the origin) to opportunities of type $k$, from the perspective of person type $p$. As outlined previously, the measure reflects the attractiveness of opportunities $W$ at location $j$ of type $k$. Finally, the term $c_{ij}$ refers to a function of travel cost between $i$ and $j$, as experienced by person $p$. -->

formulation in Eq. \ref{eq:location-based1} incorporates: (1) individual’s profiles; 2) how different profiles perceive or experience the space; and 3) constraints that may prevent segments of the population to access some type of activities. Some of these individual aspects involved in the construction of accessibility measures have been found to be relevant in empirical studies. For instance, gender in @Osland2010a or age and gender in @Theriault2013 in the study of the property prices.

Equation \ref{eq:access-general} shows the formulation potential accessibility to employment from location $A_i$, which the frequently-used in literature.

\begin{equation}
  A_{i} = \sum^N_{i=1} E_j f(c_{ij})
  \label{eq:access-general}
\end{equation}

The attractiveness of a destination is given by the number of employment opportunities at $j$, $E_j$. $f(c_{ij})$ is the impedance affecting attractiveness as a function of travel cost between $i$ and $j$. Frequently, travel cost is represented by the modelled travel time.

In Eq. \ref{eq:eq:access-general}, households and opportunities are assumed to be homogeneous, i.e., all types of employment are equally attractive for all types of residents. Following contributions from theory [@Geurs2004; @Levinson2020], an extended accessibility measure should take into consideration the characteristics of the demand (residents at origins) and supply at potential destinations. In the context of the labour market, this idea is supported by the Human Capital Theory [@Schultz1961; @Becker1994]. 

There are different versions of its operationalization in existing empirical studies [e.g. @Pereira2019b]. For instance, these characteristics can be expressed as type of person $m$ at $j$ and type of employment $m$ at $k$. Accordingly, the measure can be specified as following:

\begin{equation}
  A_{im} = \sum^N_{i=1} E_{im} f(c_{ij}),
   \label{eq:access-matching}
\end{equation}

Here, only employment opportunities $E$ matching the profile of residents between $j$ and $k$ are considered. Thus, this reflect the notion that residents would find attractive only employment opportunities that match their skills and education level with a relevant employment opportunity.

### Evaluation of public transport

Location-based accessibility measures have been used to evaluate the performance of public transport systems in policy report and research [@Ermagun2021; @Liu2022; @Conway2018; @Malekzadeh2020; @Palmateer2021]. A measure based on realistic travel times can reflect important elements of a public transport system. This is because travel time by public transport is affected by several aspects of the system, such as:

* The topology of the network, e.g. connectivity, centrality.
* Capacity of the infrastructure, e.g. degree of right of way segregation, capacity and design of stations.
* Spatial coverage, e.g. availability of infrastructure and catchment areas.
* Operational characteristics, e.g. frequency of services, overlapping routes, express services, transfer wait times.

Despite the growing relevance of accessibility measures in the evaluation of public transport infrastructure, knowledge of the implications of overlooking matching employment are limited in literature.


# Methods

## Accessibility measures examined

The general accessibility is computed as following for this work:

\begin{equation}
  A^{\text{PT}}_{jt} = \sum^J_{j=1} E_k f(d_{jkt})
   \label{eq:empirical-access-general}
\end{equation}

The implementation of matching accessibility is adapted according to the information available for the area of study and some preliminary findings. The households' education level (i.e., median number of years of education) is employed assuming correspondence with the level of salary paid in the employment market. This is further supported by the positive relationship found between education level and wages in empirical studies in the context of Mexico [@Nunez2017; @Rojas2000].

\begin{equation}
  A^{\text{PT}}_{jtm} = \sum^J_{j=1} E_{km} f(d_{jkt}),
   \label{eq:empirical-access-matching}
\end{equation}

Here, $m$ at $j$ is categorized according to the median education level of its residents by quintile and $m$ at $k$ is classified based on the average paid salary by quintile. Only employment opportunities $E$ matching the education level quantile between $j$ and $k$ are considered (when $j \ne k$).^[An exception occurs when $j = k$. In this case, employment opportunities in the origin are considered as available. This is to avoid the assumption implying that some zones have accessibility values equal to 0] This assumes that residents would deem valuable only employment opportunities that match their education level, i.e. those in the same quintile. The data sources and specific procedures to process the data are described in the next sub-section.

## Area of study

Greater Mexico City (GMC) (*Zona Metropolitana del Valle de México*) is located in in the south-centre of Mexico. In 2015, GMC had nearly 21 million inhabitants [@INEGI2015]. The official delimitation consists of 76 adjacent municipalities, which are part of three different states, namely: (1) *Ciudad de México* (16 municipalities); (2) *Hidalgo* (1 municipality), and (3) *México* (59 municipalities). The total GMC area is 7,889 km^2^, and it is home to 17% of the national population.

<!-- GMC is the main economic hub, generating a quarter of the national GDP [@OECD2015]. An overview of the composition employees by sector, utilising the North American Industry Classification System (NAICS), shows that _Retail trade_ accounted for 21% of the share in 2015 [@INEGI2015b]. This is followed by _Administrative and support and waste management and remediation services_ and _Manufacturing_ with about 14% each. Other relevant sectors are _Accommodation and food services_ (7%) and _Finance and insurance_ (6%).  -->

### The main public transport network (MPTN)

<!-- From the 34.6 million journeys completed in 2017, 64% were on foot or bicycle in at least one stage of a journey; 16% on bus rapid transit (BRT) or rail (5.4 million);^[Rail modes include metro, light rail transit (LRT), regional or commuter rail; and aerial tramway.] 36% bus or trolley; 20% on private car or motorcycle; 6% taxi; and about one percent utilised special services or other forms of transport. -->

The focus of this work is on *rapid transit* and *semi-rapid transit* infrastructure, excluding specialized modes i.e. *aerial tramway*. The combination of public transport modes included in the analyses is referred to as the *main public transport network* (MPTN). Figure \ref{fig:metro-map} shows the MPTN in GMC for the year 2019. This consists of a variety of modes operated by the following main agencies: The metro system, directly operated by the state-owned company *Sistema de Transporte Colectivo* (STC) and designated as Metro; two BRT systems, namely *Metrobus* (MB) in Ciudad de México and Mexibus (MXB) in the state of México; One LRT line operated by STE; and one regional rail line designated as _Sub_.

\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth]{figures/metro_map.png}
  \caption{Main public transport network (MPTN). Greater Mexico City, 2019.}
  \footnotesize{\textit{Source:} Verduzco-Torres (2023) based on Marco Geoestadístico Nacional 2014 Versión 6.2 (INEGI, 2014); Delimitación de las zonas metropolitanas México 2015 (INEGI, 2018); Secretaría de Movlidad de la Ciudad de México (SEMOVI, n.d.); Sistema de Transporte Masivo y Teleférico del Estado de México (SITRAMyTEM, n.d.); OpenStreetMap (OSM, n.d.).}
  \label{fig:metro-map}
\end{figure}

<!-- Until the period of study, MB operated seven corridors which add up to 240 kilometres [@Semovi2020]. In 2019, the MB system had 1.4 million passengers on average per working day. Meanwhile, the MXB system operates three BRT corridors, namely MXB line 1 (MXB-L1), MXB line 2 (MXB-II), and MXB line 3 (MXB-III). The length covered by this system sums 57.3 kilometres [@SITRAMyTEM]. A 13-kilometre light-rail also operates in the south of GMC. Mexico's City public agency *Sistema de Transporte Colectivo* (STE) owns and operates 12 metro lines. The total length of the metro network is 218 kilometres, serving about 5 million passengers a day [@Semovi2020]. The rail system includes a regional rail (*Tren Suburbano*, SUB, henceforth). This a 27-kilometre line is frequently combined with other *rapid transit* modes. For instance, 43% and 21% of the journeys that use the SUB are combined with metro or BRT, respectively. -->


The period of study was divided into seven temporal stages to reflect significant changes in the MPTN over time. Initially, changes were identified by the physical alterations in the MPTN, such as the opening of a new corridor or the extension of an existing one. Table \ref{tab:stages-network} details these changes and their corresponding temporal stages from 2010 to 2019. The stages were grouped based on the time interval between changes, with modifications occurring within six months being grouped together in the same stage. This criteria follows previous empirical studies and assumes that accessibility benefits are perceived as simultaneous when changes happen within short timeframes [@Ahlfeldt2013; @Gibbons2005].


```{r stages-network}
# Table: Network stages
stages <- read_csv("tables/transport_network_history.csv")
stages <- stages %>% 
  mutate(operation = lubridate::dmy(operation)) %>% 
  filter(operation >= lubridate::dmy("01/12/2008")) %>% 
  select(route_short_name, agency, stage_from, stage_to, operation:based_on_gtfs)
# Fix dates
stages$based_on_gtfs[1] <- "20190109"
stages$based_on_gtfs <- as.Date(stages$based_on_gtfs, format = "%Y%m%d")
# Fix stage names
stages <- stages %>% 
  mutate(period_2  = ifelse(
  grepl("^[0-9]",period_2),
  gsub("-", " to ", period_2),
  str_to_sentence(gsub("-", " ", period_2))))
stages[nrow(stages), "time_diff"] <- NA
stages$stage_to <- sub("/", " ", stages$stage_to)
stages <- select(stages, period_2, everything())
# Stages caption
stages_capt <- "Main public transport network extensions between 2010 and 2019 in Greater Mexico City grouped by temporal stage."
# Name columns
stages_headers <- c("Temporal stage", "Corridor", "Agency", "From station", "To station", "Opening date", 
                    "Previous change*", "GTFS source")
# Group rows by stage
stages_row_groups <- 
c("Stage 7: 2018 to 2019" = 1, "Stage 6: 2016 to 2018" = 1, "Stage 5: 2015 to 2016" = 1, 
  "Stage 4: 2013 to 2015" = 2, "Stage 3: 2012 to 2013" = 2, "Stage 2: 2010 to 2012" = 2, 
  "Stage 1: Before 2010" =1)
# Stages footnote
stages_footnote <- c("*Months relative to the previous modification of the main public transport network.")

# Print table
stages[,-1] %>% 
  kable(
    format = "latex",
    booktabs = TRUE,
    col.names = stages_headers[-1],
    align = "l",
    caption = stages_capt
   ) %>%
  column_spec(3:5, width = "8em") %>% 
  column_spec(6, width = "4em") %>% 
  kable_styling(font_size = 9) %>%
  pack_rows(index = stages_row_groups, italic = TRUE, bold = TRUE) %>% 
  footnote(general = stages_footnote) 
```


The MPTN stages and corresponding corresponding corridors are in Figure \ref{fig:evolution-network}.

\begin{figure}
  \centering
  \includegraphics[width=0.95\textwidth]{figures/evo_transit_network.png}
  \caption{Temporal stages of the main public transport network. Greater Mexico City, 2010 to 2019.}
  \footnotesize{\textit{Source:} The author based on Semovi (n.d.) and SITRAMyTEM (n.d.).}
  \label{fig:evolution-network}
\end{figure}

## Data


### Demographics

The demographic data employed to account for person components in accessibility measures come from the 2010 Population Census [@INEGI2012]. These data were selected for their detailed disaggregation at the city block level, including key fields such as total population and average years of education completed by individuals aged 15 and older.

The count data included were transferred to the SAS using the *areal weighting method* [@Saporito2007]. This reassigns the data according to the proportional overlay of the source and target areas. The average education level was assigned using the *population weighting method*. In addition, this takes into consideration the number of inhabitants and their type.

<!-- Figure \ref{fig:map-edulevel} displays the median years of education by quintile across SASs. The spatial patterns are consistent across units, with highly educated residents concentrated in the southwest, while lower-educated residents are mainly in peripheral areas, particularly in the southeast and north. -->

<!-- \begin{figure} -->
<!--   \centering -->
<!--   \includegraphics[width=0.9\textwidth]{data/01_census_access/edu_decile_map.png} -->
<!--   \caption{Level of education aggregated at various spatial analytical schemes. Greater Mexico City, 2010.} -->
<!--   \footnotesize{\textit{Source:} The author based on Censo de Población y Vivienda 2010 (INEGI, n.d.)} -->
<!--   \label{fig:map-edulevel} -->
<!-- \end{figure} -->

### Employment

Employment data are sourced from the 2014 Economic Census (*Censos Económicos 2014*) [@INEGI2015] and accessed at the city block level via the Microdata Lab and Remote Processing service (<https://www.inegi.org.mx/microdatos/>). The dataset includes: 1. Number of establishments (fixed locations producing goods or services); 2. Employed personnel (individuals active in establishments, including volunteers and family members), referred to as ‘employment’; and 3. Total gross remuneration paid for the relevant year. 

Employment information which came aggregated at higher geographic level than the city-block due to privacy (<8%) were re-allocated using proportional estimates based on the spatial distribution from the 2015 DENUE [@INEGI2015a]. This is an open-access, point-level dataset of individual establishments. The employment data was assigned from city blocks to SASs following the same approach employed for the demographic data, i.e. according to the areal weighting method.

The average income was computed for each areal unit based on the total number of employment and the total gross remuneration paid.


### Public transport itineraries

The estimated travel time used for accessibility measures is based on scheduled public transport services. The Transport Authority in Mexico City (*Secretaría de Movilidad de la Ciudad de México*, Semovi) released the first *general transit feed specification* (GTFS) files in 2013, which were accessed via the Open Mobility website (<https://openmobilitydata.org/>). These datasets cover the agencies operating most MPTN services in GMC. However, there are some limitations: firstly, the geographic coverage primarily includes services within Ciudad de México, excluding BRT services provided by MXB. Secondly, the temporal coverage starts in 2013, with updates occurring only sporadically, approximately once a year.

To overcome the first limitation, the geographic coverage of the official GTFS data was augmented to include three BRT corridors operated by MXB. This was carried out *manually* based on unstructured information provided by the Transport Agency of the State of Mexico (*Sistema de Transporte Masivo y Teleferico del Estado de México*, Sitramytem). This corresponded to services for 2019 only.

Regarding the second limitation, the MPTN's temporal stages were reconstructed using the GTFS standard. The primary input for each stage was the closest available official GTFS data reflecting changes in the MPTN, with only the relevant services for each corridor in their respective stages being retained. For MXB services, 2019 data was considered due to the lack of earlier details, assuming uniform characteristics of the BRT corridors across the stages. Similarly, for other corridors included in early stages, services were assumed to reflect 2013 levels due to the unavailability of earlier data. All non-MPTN services were excluded, thereby focusing exclusively on the benefits of the MPTN.

### Road network and pedestrian infrastructure

The road and pedestrian infrastructure network is sourced from *OpenStreetMap* (OSM) (Manually downloaded on 30th October 2019 through a web export tool available in <https://export.hotosm.org/es/v3/>]). This enables multi-modal journeys to realistically represent the access/egress stage to a public transport service when combining walking and public transport.

## Travel costs and impedance parameters

### Modeled travel time

Travel time matrices were calculated for each origin-destination (OD) pair across SASs using a combination of walking and public transport within the main public transport network. OD points $i$ and $j$ are represented by population-weighted centroids of spatial units containing economic activity (from the 2014 Economic Census) or demographic data (from the 2010 Population Census). Travel times are capped at 180 minutes, based on patterns observed in the 2017 Travel Survey.

Travel times were modelled using a door-to-door approach with OpenTripPlanner (OTP) version '1.4.0' [@OTP] via the implementation of an open-access routine written `Python` [@Pereira2019a]. This approach considers access, waiting, onboard, transfer, and egress times. All-to-all travel time matrices were generated for each SAS across seven temporal stages using reconstructed GTFS files and road/pedestrian networks, with eight random departure times between 10 a.m. and 2 p.m. The median travel time was used for accessibility models [@Conway2017]. This process, computationally intensive due to the high resolution of SASs, resulted in 2.1 billion routes. Matrices were then merged, and median travel times calculated, with walking distances limited to 3 km when using public transport and 4.8 km (60 minutes) for walking-only trips.

The internal-zone travel-times were estimated based on the surface area as following: $d_{i=j} = \frac{1}{2} \sqrt{ \frac{area_i}{\pi}}$ [@Frost1995; @Stepniak2015]. The metric distance was transformed to travel time using the average speed modelled for short journeys (less than 3 km).

### Spatial Decaying Function and Calibration of Accessibility Parameters

The spatial decaying function and parameters for accessibility measures were determined using estimates from a trip distribution model, following @Handy1997. This model is used to understand how commuters in Greater Mexico City (GMC) discount employment opportunities based on travel time. The analyses use data from the 2017 Travel Survey, focusing on weekday public transport and walking trips.

A series of alternative specifications were estimated within a doubly constrained gravity model framework, expressed as following:

\begin{equation}
  T_{jk} = A_j O_j B_k D_k f(d_{jk})
  \label{eq:gravity-model}
\end{equation}

where
\begin{equation} 
  A_j = \left[ \sum{B_i D_i f(d_{jk})} \right]^{-1}
\end{equation}

and
\begin{equation} 
  B_k = \left[ \sum{A_k O_k f(d_{jk}) }  \right] ^{-1}
\end{equation}

with balancing factors $A_j$ and $B_k$ calculated iteratively. The model estimates commuter flows ($T_{jk}$) between origins ($j$) and destinations ($k$), with constraints on the number of journeys at each. $f(d_{jk})$ is a spatial deterrence or decaying function. The inverse power ($d_{jk}^{-\beta}$) and negative exponential ($e^{(-\beta d_{jk})}$) decaying functions were tested. $\beta$ represents the deterrence parameter.

A modified specification intended to capture an add-up cost for inter-zone journeys or analogously a benefit of residing and working within the same zone is considered [@Thorsen1998]. This includes a parameter $\mu$ which affects only the diagonal elements of the OD matrix. The term is given by the following expression:

\begin{equation} 
  \mu \delta_{jk}
\end{equation}

where Kronecker delta takes the following value:

\begin{equation}
  \delta_{ij} \left\{
    \begin{array}{ll}
        0 & \quad \text{if   } j \neq k \\
        1 & \quad \text{if   } j = k \\
    \end{array} 
   \right.\\
\end{equation}

The simultaneous calibration of the model parameters $\beta$ and $\mu$ is achieved via optimization by minimizing the standardized root mean square error (SRMSE) using the *Nelder-Mead's simplex algorithm* as implemented in base `R` [@R-base].

The preliminary analyses shows that the negative exponential function provided more accurate estimates compared to the inverse power. Also, the use of the within-zone parameter $\mu$ improves the estimates. Therefore, the selected specification is the modified negative exponential including $\mu$. The parameters employed to estimate accessibility are shown in Table \label{tab:access-pars}.

# Results

\FloatBarrier
# Conclusions

\FloatBarrier
# References